Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-06-01 Ben_Manuscripts/hdphmm/Draft.tex
added TOC graphic
comments

  In this work, we are particularly interested in the design of lyotropic liquid 
  crystal (LLC) monomers, a class of amphiphilic molecules whose ordered self-assembled
  phases can be cross-linked into mechanically strong membranes capable of highly 
  selective separations. Inverted hexagonal phase, or H\textsubscript{II}, LLC membranes
  are characterized by hexagonally packed, uniform-sized and straight pores, an ideal
  geometry for high throughput transport. Their pores are lined with the LLC monomer's
  functional groups which can potentially be designed to interact with solutes in a 
  chemically-specific manner. It would be extremely useful to experimentalists if we 
  could connect LLC monomer design with measurable quantities. This goal is complicated
  by the non-Brownian hopping and trapping behavior induced by the membrane's diverse
  topology and inhomogeneous structure resulting in subdiffusive behavior out to 
  multiple microseconds of simulation.~\cite{coscia_understanding_2019,coscia_chemically_2019}

  We have already advanced our understanding of LLC monomer design in our previous work
  by building stochastic time series models based on the chemical intuition gained
  from qualitative MD studies of solute transport in an H\textsubscript{II} phase LLC
  membrane.~\cite{coscia_chemically_2019,coscia_capturing_2020} We identified three 
  classes of trapping behavior which result in subdiffusion out to the microsecond 
  time scale. In our first approach to stochastically model this behavior, we treated 
  the system's dynamics as a sequence of anti-correlated hops between power law distributed periods of 
  entrapment, a framework called subordinated fractional Brownian or L\'evy
  motion.~\cite{thiel_weak_2014,teuerle_modeling_2013} In a second approach, we treated 
  solute motion as a Markov state model~\cite{pande_everything_2010} with state-dependent
  dynamics, where we parameterized the state transition probabilities between each of
  eight discrete states, defined by the observed trapping mechanisms, as well as the 
  solute dynamics within each of these states. We showed how one could use realizations
  of any stochastic model like these in order to predict macroscopic flux and selectivity.

@@

  By placing a hierarchical Dirichlet process (HDP) prior on the observed time
  series data, it is possible to extend HMMs to an unknown number of hidden
  states, thus allowing the model's complexity to be driven by the complexity
  of the data.~\cite{teh_hierarchical_2006} We will refer a model calibrated with
  this type of prior as an HDP-HMM; this model class is sometimes referred to
  as an infinite hidden Markov model.~\cite{beal_infinite_2002} The HDP-HMM 
  elegantly overcomes the main shortcoming of standard HMMs, the need to choose
  the number of states \textit{a priori}, making its use more advantageous in
  a diversity of applications including economic data and single particle trajectory 
  analysis.~\cite{shi_identifying_2016,hines_analyzing_2015,sgouralis_introduction_2017,jazani_alternative_2019,wang_single-molecule_2014,persson_extracting_2013}

@@

  The utility of the sticky HDP-HMM has been further extended to account for
  inherent temporal dependence between measurements common to many time series
  datasets. For example, in both MD and single particle tracking, there is 
  strong temporal dependence/correlation between successive measurements even
  within a fixed state.~\cite{calderon_data-driven_2014}
  Fox et al. formulated a method that accounts for temporal dependence betwee
  successive observations by incorporating Markov jump linear systems (MJLSs)
  into infinite state models.~\cite{fox_nonparametric_2009} An MJLS models 
  each dynamical mode as a linear dynamical process. Two MJLS sub-classes are 
  switching linear dynamical systems (SLDSs) and switching vector autoregressive
  (VAR) processes. Switching VAR processes are a sub-class of SLDSs that do not
  incorporate measurement noise. Since MD simulations do not contain measurement
  noise, we will model the dynamical modes exhibited by MD solute trajectories 
  as VAR processes and we will refer to our model as an application of the 
  sticky HDP-AR-HMM in order to maintain its distinction from the even more
  general sticky HDP-SLDS models.

  To our knowledge, applying the HDP-AR-HMM in order to gain kinetic and
  mechanistic insight from MD simulation trajectories has not been done
  before. Typical applications of the HDP-AR-HMM center around determining the 
  number of states, the state sequence and analysis of the emission 
  distributions of a time series. The addition of temporal correlation to 
  the inference procedure has primarily been used to improve state sequence 
  identification.~\cite{calderon_inferring_2015} In our approach, we use the
  HDP-AR-HMM not only to segment a collection of MD time series into distinct 
  dynamical modes, but to uncover unknown transport mechanisms and to create 
  improved predictive models. We cluster the parameters of individual states
  in order to reduce the total state space down to an interpretable size
  This allows us to learn the implied dominant physical mechanisms and describe
  solute kinetics based on interpretation of distinct parameters that dominate
  the ensemble of MD trajectories. We can use the model parameters determined 
  either before or after clustering to generate new time series trajectories 
  which enable projections of the MD-generated time series to arbitrary length
  time scales.

@@

  \item Second, we want to know if the HDP-AR-HMM algorithm can help
  efficiently uncover underlying transport mechanisms which give rise to different 
  solute dynamical behavior. We cluster similar state parameters across all
  trajectories in order to reduce the total state space and to understand which
  segments of the solute trajectories exhibit similar behavior. We use the 
  parameters of the clustered states identified by the HDP-AR-HMM in order to 
  infer dominant solute-membrane interactions and transport mechanisms. We 
  support our mechanistic hypotheses via quantitative comparison to various 
  physical membrane properties and solute-membrane interactions.

@@

  We studied transport of solutes in the H\textsubscript{II} phase using an
  atomistic molecular model of four pores in a monoclinic unit cell with 
  10\% water by weight. Approximately one third of the water molecules 
  occupy the tail region with the rest are near the pore center.~\cite{coscia_chemically_2019}

@@

  We use the sticky HDP-AR-HMM in order to assign the time-correlated positional 
  fluctuations of solute center-of-mass MD trajectories to discrete dynamical modes while 
  simultaneously inferring the VAR parameters of those modes. We use the sticky model
  in order to encourage higher self-transition probabilities since we physically expect
  solutes to stay trapped in states for appreciable amounts of time.

@@

  We clustered states with like parameter sets in order to reduce the state space to
  a more easily interpretable size 
  % CPC: perhaps keep hammering home not only reducing, but also reemphasize why absolute
  % value of some parameters really not of interest 
  %MRS: explain why parameters of single states don't give insight; more context needed?  Is CPC talking about
  %MRS: some some of the parmameters not being interesting in general, not just the values for some states?
  and because the parameters of a single state alone does not always have enough context
  to provide meaningful physical insight. For each solute studied, we identified 
  approximately 300-500 independent states, each with separate VAR(1) parameters.
  Many of these states exhibit very similar dynamical behavior except their mean 
  levels in each dimension are different.

@@

  We obtained estimates of clustered $A$ and $\Sigma$ parameters as well as a
  consensus $T$ matrices, describing transitions between all possible clustered states 
  exhibited by the ensemble of trajectories, using the inference component of 
  the HDP-AR-HMM algorithm. We applied the HDP-AR-HMM simultaneously to all 24
  zeroed solute trajectories with state sequences fixed according to the remapped 
  clustered state assignments. Simultaneous application of the HDP-AR-HMM to 
  all trajectories as we have done is only possible because the state sequence 
  is fixed. In cases where one does not know the state sequence and would like to 
  incorporate data from multiple trajectories, one can apply the beta process (BP)
  AR-HMM.~\cite{fox_sharing_2009} However the data requirements may be quite high
  for a good parameterization, which is why we elected to follow the clustering 
  approach described thus far.
  %BJC: I actually found 3 different ways do this. All giving the same results.
  % Method 1: described above - this is the most efficient way
  % Method 2: Link all 24 trajectories by adding a coffin state (basically a 
  % state that is a vector of zeros) between each traj
  % Method 3: Parameterize each trajectory independently, average together. T is
  % estimated by making a count matrix based on inputs from each trajectory

@@

  We reported the mean and $1 \sigma$ confidence intervals of the MSDs based on the 
  results of 200 bootstrap trials. Our bootstrapping procedure varied dependent on the
  method of trajectory generation:
  \begin{enumerate}[label={Method \theenumi :}, leftmargin=3.5\parindent]
  	\item For each bootstrap trial, we generated 24 stochastic trajectories, one from
  	each of the 24 independent HDP-AR-HMM models. The result of the trial is the mean
  	of the MSDs calculated from each trajectory constituting the ensemble. 
    \item We generated an ensemble of 1000 independent stochastic trajectories using
    the clustered parameter sets. For each trial, we calculated the MSD of 1000 
    trajectories sampled, with replacement, from the ensemble.
  \end{enumerate}

@@

  % BJC: Note that this figure is not finalized. It just takes a while to make 
  % because of bootstrapping and I keep forgetting to let it run overnight. I 
  % actually did it once, but had set save=False in the function I used in a jupyter notebook
  % In its final form, the errorbars will get smaller, and the means don't change by much.
  \begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{selectivity.pdf}
  \caption{Of the solutes studied, the current generation of H\textsubscript{II}
  LLC membranes may have the most success selectively removing acetic acid and,
  to a lesser extent, urea from solution containing either ethylene glycol or 
  methanol.
  }\label{fig:selectivity}
  \end{figure}

@@

     \item This state has intermediate dwell times with high past-hop dependence (high
     entries of $A$). The cluster's radial mean is 1.5 nm which is about halfway between
     the pore center and tail ends. Evidence from Figure~\ref{fig:clustered_traj_MET} 
     suggests that this state occurs essentially evenly across possible radial
     distances from the pore center.

@@

  %\newpage

  %\section*{TOC Graphic}

@@

% LocalWords:  BJC micropollutants Desalination permeability nanofiltration LLC
% LocalWords:  Amphiphilic nanostructures Lyotropic amphiphilic lyotropic MSDDM
% LocalWords:  solutes selectivities solute nanoscopic timescales MSDs MSD HMMs
% LocalWords:  nonparameteric Parrinello Rahman barostat rescale transitioning
% LocalWords:  al's HMM Dirichlet HDP DP Wishart TODO SI al MATLAB teh gael pdf
% LocalWords:  nonparametric bayesian solute's reparameterized intracluster ca
% LocalWords:  delocalize Luzar Chandler histogrammed interpolator methanol's
% LocalWords:  nanostructure carboxylate VMD nonidealities hbonds outlier GAANN
% LocalWords:  autogregressive PRF ACS XSEDE ACI HDP-AR-HMM nanoporous flowback evy
% LocalWords:  timescale subdiffusive coscia subdiffusion calderon GROMACS der
% LocalWords:  berendsen gromacs spoel hess emailed plateauing hmm ne gelman AR
% LocalWords:  walt numpy GCL URE ACH IMMM subsubsection modT convective vmd Na
% LocalWords:  scannable subsubcaptions errorbars uncharacteristicness facelift
% LocalWords:  PSC RMACC hdphmm TOC nominalizations dischinger ramaseshan Baum
% LocalWords:  functionalized teuerle jasra markov bioinformatics hughes yoon
% LocalWords:  hines diarization MJLSs MJLS SLDSs SLDS hamada jk params MNIW
% LocalWords:  plateaued nclusters leftmargin trans diffusivity sorption alkane
% LocalWords:  ness substituents desalination fonseca couto welch singhal noe
% LocalWords:  shirtsgroup superdiffusive diffusivities

>>>>

  In this work, we are particularly interested in the design of lyotropic liquid 
  crystal (LLC) monomers, a class of amphiphilic molecules whose ordered self-assembled
  phases can be cross-linked into mechanically strong membranes capable of highly 
  selective separations. Inverted hexagonal phase, or H\textsubscript{II}, LLC membranes
  are characterized by hexagonally packed, uniform-sized and straight pores, an ideal
  geometry for high throughput transport. Their pores are lined with the LLC monomer's
  functional groups which can potentially be designed to interact with solutes in a 
  chemically-specific manner. It would be extremely useful to experimentalists if we 
  could connect LLC monomer design with measurable quantities. This goal is complicated
  by the non-Brownian hopping and trapping behavior induced by the membrane's diverse
  topology and inhomogeneous structure resulting in subdiffusive behavior out to 
  multiple microseconds of simulation.~\cite{coscia_understanding_2019,coscia_chemically_2019}
  %MRS: citation missing for coscia_chemically_2019
  We have already advanced our understanding of LLC monomer design in our previous work
  by building stochastic time series models based on the chemical intuition gained
  from qualitative MD studies of solute transport in an H\textsubscript{II} phase LLC
  membrane.~\cite{coscia_chemically_2019,coscia_capturing_2020} We identified three 
  classes of trapping behavior which result in subdiffusion out to the microsecond 
  time scale. In our first approach to stochastically model this behavior, we treated 
  the system's dynamics as a sequence of anti-correlated hops between power law distributed periods of 
  entrapment, a framework called subordinated fractional Brownian or L\'evy
  motion.~\cite{thiel_weak_2014,teuerle_modeling_2013} In a second approach, we treated 
  solute motion as a Markov state model~\cite{pande_everything_2010} with state-dependent
  dynamics, where we parameterized the state transition probabilities between each of
  eight discrete states, defined by the observed trapping mechanisms, as well as the 
  solute dynamics within each of these states. We showed how one could use realizations
  of any stochastic model like these in order to predict macroscopic flux and selectivity.

@@

  By placing a hierarchical Dirichlet process (HDP) prior on the observed time
  series data, it is possible to extend HMMs to an unknown number of hidden
  states, thus allowing the model's complexity to be driven by the complexity
  of the data.~\cite{teh_hierarchical_2006} We will refer to a model calibrated with
  this type of prior as an HDP-HMM; this model class is sometimes referred to
  as an infinite hidden Markov model.~\cite{beal_infinite_2002} The HDP-HMM 
  elegantly overcomes the main shortcoming of standard HMMs, the need to choose
  the number of states \textit{a priori}, making its use more advantageous in
  a diversity of applications including economic data and single particle trajectory 
  analysis.~\cite{shi_identifying_2016,hines_analyzing_2015,sgouralis_introduction_2017,jazani_alternative_2019,wang_single-molecule_2014,persson_extracting_2013}

@@

  The utility of the sticky HDP-HMM has been further extended to account for
  inherent temporal dependence between measurements common to many time series
  datasets. For example, in both MD and single particle tracking, there is 
  strong temporal dependence/correlation between successive measurements even
  within a fixed state.~\cite{calderon_data-driven_2014}
  Fox et al. formulated a method that accounts for temporal dependence between
  successive observations by incorporating Markov jump linear systems (MJLSs)
  into infinite state models.~\cite{fox_nonparametric_2009} An MJLS models 
  each dynamical mode as a linear dynamical process. Two MJLS sub-classes are 
  switching linear dynamical systems (SLDSs) and switching vector autoregressive
  (VAR) processes. Switching VAR processes are a sub-class of SLDSs that do not
  incorporate measurement noise. Since MD simulations do not contain measurement
  noise, we will model the dynamical modes exhibited by MD solute trajectories 
  as VAR processes and we will refer to our model as an application of the 
  sticky HDP-AR-HMM in order to maintain its distinction from the even more
  general sticky HDP-SLDS models.

  To our knowledge, applying the HDP-AR-HMM in order to gain kinetic and
  mechanistic insight from MD simulation trajectories has not been done
  before. Typical applications of the HDP-AR-HMM center around determining the 
  number of states, the state sequence and analysis of the emission 
  distributions of a time series. The addition of temporal correlation to 
  the inference procedure has primarily been used to improve state sequence 
  identification.~\cite{calderon_inferring_2015} In our approach, we use the
  HDP-AR-HMM not only to segment a collection of MD time series into distinct 
  dynamical modes, but to uncover unknown transport mechanisms and to create 
  improved predictive models. We cluster the parameters of individual states
  in order to reduce the total state space down to an interpretable size.
  This allows us to learn the implied dominant physical mechanisms and describe
  solute kinetics based on interpretation of distinct parameters that dominate
  the ensemble of MD trajectories. We can use the model parameters determined 
  either before or after clustering to generate new time series trajectories 
  which enable projections of the MD-generated time series to arbitrary length
  time scales.

@@

  \item Second, we want to know if the HDP-AR-HMM algorithm can help
  efficiently uncover underlying transport mechanisms which give rise to different 
  solute dynamical behavior. We cluster 
  %MRS:
  %similar state parameters 
  states with similar parameters
  across all
  trajectories in order to reduce the total state space and to 
  %MRS:
  identify
  %understand 
  which
  segments of the solute trajectories exhibit similar behavior. We use the 
  parameters of the clustered states identified by the HDP-AR-HMM in order to 
  infer dominant solute-membrane interactions and transport mechanisms. We 
  support our mechanistic hypotheses via quantitative comparison to various 
  physical membrane properties and solute-membrane interactions.

@@

  We studied transport of solutes in the H\textsubscript{II} phase using an
  atomistic molecular model of four pores in a monoclinic unit cell with 
  10\% water by weight. Approximately one third of the water molecules 
  occupy the tail region with the rest are near the pore center.~\cite{coscia_chemically_2019}
  %MRS: fix citation above.

@@

  We use the sticky HDP-AR-HMM in order to assign the time-correlated positional 
  fluctuations of solute center-of-mass MD trajectories to discrete dynamical modes while 
  simultaneously inferring the VAR parameters of those modes. We use the sticky model
  in order to encourage higher self-transition probabilities since we physically expect
  %MRS: maybe ``physically observe?'' instead of physically expect.
  solutes to stay trapped in states for appreciable amounts of time.

@@

  We clustered states with like parameter sets in order to reduce the state space to
  a more easily interpretable size 
  % CPC: perhaps keep hammering home not only reducing, but also reemphasize why absolute
  % value of some parameters really not of interest 
  %MRS: explain why parameters of single states don't give insight; more context needed?  Is CPC talking about
  %MRS: some some of the parameters not being interesting in general, not just the values for some states?
  and because the parameters of a single state alone does not always have enough context
  to provide meaningful physical insight. For each solute studied, we identified 
  approximately 300-500 independent states, each with separate VAR(1) parameters.
  Many of these states exhibit very similar dynamical behavior except their mean 
  levels in each dimension are different.

@@

  We obtained estimates of clustered $A$ and $\Sigma$ parameters as well as a
  %MRS: check again agreement on $T$ and $P$ - you switched to $P$, right? Or are they 2 different things?
  consensus $T$ matrices, describing transitions between all possible clustered states 
  %MRS: do you explicitly say how to get the consensus transition matrices?
  exhibited by the ensemble of trajectories, using the inference component of 
  the HDP-AR-HMM algorithm. We applied the HDP-AR-HMM simultaneously to all 24
  zeroed solute trajectories with state sequences fixed according to the remapped 
  clustered state assignments. Simultaneous application of the HDP-AR-HMM to 
  all trajectories as we have done is only possible because the state sequence 
  is fixed. In cases where one does not know the state sequence and would like to 
  incorporate data from multiple trajectories, one can apply the beta process (BP)
  AR-HMM.~\cite{fox_sharing_2009} However the data requirements may be quite high
  for a good parameterization, which is why we elected to follow the clustering 
  approach described thus far.
  %BJC: I actually found 3 different ways do this. All giving the same results.
  % Method 1: described above - this is the most efficient way
  % Method 2: Link all 24 trajectories by adding a coffin state (basically a 
  % state that is a vector of zeros) between each traj
  % Method 3: Parameterize each trajectory independently, average together. T is
  % estimated by making a count matrix based on inputs from each trajectory

@@

  We reported the mean and $1 \sigma$ confidence intervals of the MSDs based on the 
  results of 200 bootstrap trials. Our bootstrapping procedure varied dependent on the
  method of trajectory generation:
  %MRS: this is a bit confusing.  Is there an expectation these should give consistent parameters?
  %MRS: the second is boostrapping an MSD over 1000 trajectories, and the first is boostrapping an MSD over 24 trajectories.
  %MRS: these should have different amounts of noise, making them hard to compare? Is this an issue?
  \begin{enumerate}[label={Method \theenumi :}, leftmargin=3.5\parindent]
  	\item For each bootstrap trial, we generated 24 stochastic trajectories, one from
  	each of the 24 independent HDP-AR-HMM models. The result of the trial is the mean
  	of the MSDs calculated from each trajectory constituting the ensemble. 
    \item We generated an ensemble of 1000 independent stochastic trajectories using
    the clustered parameter sets. For each trial, we calculated the MSD of 1000 
    trajectories sampled, with replacement, from the ensemble.
  \end{enumerate}

@@

  % BJC: Note that this figure is not finalized. It just takes a while to make 
  % because of bootstrapping and I keep forgetting to let it run overnight. I 
  % actually did it once, but had set save=False in the function I used in a jupyter notebook
  % In its final form, the errorbars will get smaller, and the means don't change by much.
  \begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{selectivity.pdf}
  \caption{Of the solutes studied, 
  %MRS: add qualifier.
  the models predict that 
  the current generation of H\textsubscript{II}
  LLC membranes may have the most success selectively removing acetic acid and,
  to a lesser extent, urea from solution containing either ethylene glycol or 
  methanol.
  }\label{fig:selectivity}
  \end{figure}

@@

     \item This state has intermediate dwell times with high past-hop dependence (large
     entries of $A$). The cluster's radial mean is 1.5 nm which is about halfway between
     the pore center and tail ends. Evidence from Figure~\ref{fig:clustered_traj_MET} 
     suggests that this state occurs essentially evenly across possible radial
     distances from the pore center.

@@

  \newpage

  \section*{TOC Graphic}

  \begin{figure}[!htb]
  \centering
  \includegraphics[width=3.25in]{hdphmm_TOC.pdf}
  \end{figure}

@@

% LocalWords:  BJC micropollutants Desalination permeability nanofiltration LLC
% LocalWords:  Amphiphilic nanostructures Lyotropic amphiphilic lyotropic MSDDM
% LocalWords:  solutes selectivities solute nanoscopic timescales MSDs MSD HMMs
% LocalWords:  nonparameteric Parrinello Rahman barostat rescale transitioning
% LocalWords:  al's HMM Dirichlet HDP DP Wishart TODO SI al MATLAB teh gael pdf
% LocalWords:  nonparametric bayesian solute's reparameterized intracluster ca
% LocalWords:  delocalize Luzar Chandler histogrammed interpolator methanol's
% LocalWords:  nanostructure carboxylate VMD nonidealities hbonds outlier GAANN
% LocalWords:  autogregressive PRF ACS XSEDE ACI HDP-AR-HMM nanoporous flowback evy
% LocalWords:  timescale subdiffusive coscia subdiffusion calderon GROMACS der
% LocalWords:  berendsen gromacs spoel hess emailed plateauing hmm ne gelman AR
% LocalWords:  walt numpy GCL URE ACH IMMM subsubsection modT convective vmd Na
% LocalWords:  scannable subsubcaptions errorbars uncharacteristicness facelift
% LocalWords:  PSC RMACC hdphmm TOC nominalizations dischinger ramaseshan Baum
% LocalWords:  functionalized teuerle jasra markov bioinformatics hughes yoon
% LocalWords:  hines diarization MJLSs MJLS SLDSs SLDS hamada jk params MNIW BP
% LocalWords:  plateaued nclusters leftmargin trans diffusivity sorption alkane
% LocalWords:  ness substituents desalination fonseca couto welch singhal noe
% LocalWords:  shirtsgroup superdiffusive diffusivities jazani wang persson CPC
% LocalWords:  shortlag jupyter ester
